---
title: "Almond_Profit_Function"
author: "Madison Enda"
date: 04/20/2025
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

Load in the initial data and relevant packages

```{r}
library(tidyverse)
library(here)
library(janitor)
```

```{r}
climate_data <- read_table(here("data", "clim.txt")) %>% clean_names()
```

1.) Bring in the almond yield model

```{r}
#' Calculate Almond Yield
#'
#' @description Computes the maximum, minimum, and mean almond yield for each year 
#' based on climate data inputs such as minimum February temperature and January precipitation.
#'
#' @author Madison Enda and Joshua Mull for EDS230 with Naomi Tague
#' 
#' @param climate_data A data frame containing climate data with columns for `month`, `year`, 
#' `tmin_c` (minimum daily temperature in Celsius), and `precip` (daily precipitation in millimeters).
#' @param Tn1 Coefficient for the minimum February temperature.
#' Default is -0.015.
#' @param Tn2 Quadratic term coefficient of the minimum February temperature.
#' Default is -0.0046.
#' @param P1 Coefficient for January precipitation. Default is -0.07.
#' @param P2 Quadratic term coefficient of January precipitation. Default is 0.0043.
#' @param i Intercept term 0.28.
#'
#' @return A list containing: min, max and mean almond yields across all years 

almond_yield <- function(climate_data, Tn1 = -0.015, Tn2 = -0.0046, P1 = -0.07, P2 = 0.0043, i = 0.28) {
  
  # filter data for min temp in February 
  feb_temp <- climate_data %>%
    filter(month == 2) %>% 
    group_by(year) %>%
    summarize(min_feb_temp_c = min(tmin_c, na.rm = TRUE), .groups = "drop") 
  
  # filter data for total rain in January 
  jan_p <- climate_data %>%
    filter(month == 1) %>% 
    group_by(year) %>% 
    summarize(jan_precip = sum(precip, na.rm = TRUE), .groups = "drop") 
  
  # join data frames 
  climate_data_jan_feb <- left_join(feb_temp, jan_p, by = "year")
  
  # Compute yield for each year
  climate_data_jan_feb <- climate_data_jan_feb %>%
    mutate(yield = Tn1 * min_feb_temp_c + Tn2 * min_feb_temp_c^2 + P1 * jan_precip + P2 * jan_precip^2 + i)
  
  # Compute max, min, mean result
    maxyield <- max(climate_data_jan_feb$yield, na.rm = TRUE)
    minyield <- min(climate_data_jan_feb$yield, na.rm = TRUE)
    meanyield <- mean(climate_data_jan_feb$yield, na.rm = TRUE)
  
  # Return max, min, and mean yields
  return(meanyield)
}
```

Create data frame of average price per lbs of California almonds by year. Data sources from USDA and Merlo Farming Group: https://www.nass.usda.gov/Statistics_by_State/California/Publications/Specialty_and_Other_Releases/Almond/Forecast/202405almpd.pdf https://www.merlofarminggroup.com/almonds/Independence

Create an almond profit model that utilizes outputs from the almond yield model

```{r}
#' computes profit from power generation
#' @param farm_size (acres of farm land)
#' @param year (year when almonds were harvested, from 1995-2025)
#' @param meanyield (result from almond_yield function)
#' @return meanprofit 
profit_from_almond_yield <- function(farm_size, year) {

  # Create vectors of year and price
  date <- c(1995, 1996, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025)
  price <- c(2.48, 2.08, 1.56, 1.41, 0.86, 0.97, 0.91, 1.11, 1.57, 2.21, 2.81, 2.06, 1.75, 1.45, 1.65, 1.79, 1.99, 2.58, 3.21, 4.00, 3.13, 2.39, 2.53, 2.50, 2.45, 1.71, 1.86, 1.40, 1.64, 1.91, 2.60)
  
  # Join vectors in data frame
  almond_price <- data.frame(date, price)
  
  # Create a column that is price per ton
  almond_price <- almond_price %>%
  mutate(price_per_ton = price * 2000 )

 # Filter the data to year parameter
  almond_price_filtered <- almond_price %>%
    filter(date == year)
  
  # Call almond yield function to find meanyield
  meanyield <- almond_yield(climate_data)
  
  # Use the price per ton from subsequent year to find price by acres and 
  meanprofit <- meanyield * farm_size *  almond_price_filtered$price_per_ton
  
  return(meanprofit)
}
```

2.) Apply profit model to the almond model

```{r}
profit_from_almond_yield(farm_size= 50, year= 1999)
```

3.) Sensitivity analysis of the almond profit model

For this analysis we used 383 acres as the average farm size in California according to farmflavor.com found here 

[Farm Flavor](https://farmflavor.com/california/top-california-agriculture-facts/)


```{r}
date <- c(1995, 1996, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025)
price <- c(2.48, 2.08, 1.56, 1.41, 0.86, 0.97, 0.91, 1.11, 1.57, 2.21, 2.81, 2.06, 1.75, 1.45, 1.65, 1.79, 1.99, 2.58, 3.21, 4.00, 3.13, 2.39, 2.53, 2.50, 2.45, 1.71, 1.86, 1.40, 1.64, 1.91, 2.60)
  
# Join vectors in data frame
almond_price <- data.frame(date, price)
  
# Create a column that is price per ton
almond_price <- almond_price %>%
mutate(price_per_ton = price * 2000)
```

The mean value of price per ton was calculated to be 4036.77, this will be used as our mean for the sensitivity analysis of the year parameter, for which we will calculate z-scores. The farm size we are using is 383 acres, which was sited as the average farm size 

```{r}
# Find meanprofit for each year
year_variation_profit <- profit_from_almond_yield(farm_size = 383, year = date)

# Use the year and meanprofit results to create a new dataframe
year_variation_df <- data.frame(year_variation_profit, date) 

# Create column of z-score values
year_variation_df <- year_variation_df %>%
  mutate(z_score = scale(year_variation_df$year_variation_profit)) 

```

The year that we are selecting is 2011, where the price_per_ton value was closest to the mean calculated value for all years. The will run through a list of farm sizes centered at 383, with 31 values (chosen merely to match length of yearly data). 

```{r}
# Create list of farm sizes (in acres) centered around mean 383
numbers <- c(
  250, 350, 400, 500, 600, 300, 450, 390, 370, 410,
  420, 360, 430, 395, 380, 405, 470, 490, 375, 385,
  415, 440, 320, 340, 520, 310, 395, 450, 480, 370, 400
)

# Find meanprofit for each farm size
acre_variation_profit <- profit_from_almond_yield(farm_size = numbers, year = 2011)

# Create df 
acre_variation_profit_df <-  data.frame(acre_variation_profit, year = 2011) 

# Create column of z-score values
acre_variation_profit_df <- acre_variation_profit_df %>%
  mutate(z_score = scale(acre_variation_profit_df$acre_variation_profit))
```


4.) First graph of almond profit model that shows yield anomaly for each year, accounting for uncertainty in the parameters

```{r}
ggplot(year_variation_df, aes(x = date, y = z_score[,1])) +
  geom_line(color = "skyblue", size = 1) +
  labs(
    title = "Z-Score Variation Over Time",
    x = "Date",
    y = "Z-Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12),
    panel.grid.major = element_line(color = "lightgray"),
    panel.grid.minor = element_blank()
  )
```

Second graph of almond profit model that shows how yield anomaly in general varies with your parameters

```{r}
ggplot(acre_variation_profit_df, aes(x = z_score[,1])) +
  geom_histogram(aes(y = ..density..), binwidth = 0.5, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkblue", size = 1) +
  labs(
    title = "Density and Histogram of Z-Scores",
    x = "Z-Score",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12),
    panel.grid.major = element_line(color = "lightgray"),
    panel.grid.minor = element_blank()
  )
```

5.  Write a short paragraph (in a Quatro document) to summarize you interpretation of your model results (e.g what do 'take away' from your model and sensitivity analysis)

- Our model utilizes yearly averages of California almond profit from USDA data recorded from 1995 to 2024 (2025 data added from another source), and farm size was recorded from US census data. 
- Our sensitivity analysis confirms some biases in our data, as well as trends that we were anticipating. Firstly, the graph of z-scores by year shows that mean profit varies wildly with year information, indicating that factors associated with temporal trends play a large in overall almond yield and profit (such as weather events, natural disasters, changes in the state of the US economy, etc.). 
- With the second z-score distribution of farm size fluctuation, the relationship between farm size and mean profit is fairly linear (as farm size increases profit increases). However, it should be noted that the footprint of the majority of these California farms is rather small. We attempted to randomly generate a series of 31 values centered at 383, in order to decrease the affect of outliers (in reality the largest California Almond farm is over 100,000 acres) but we were still able to show this trend on a smaller scale, as most values fall below the mean, with only one or two high above. 
